`flippant` --- An `R` package for the automated analysis of fluorescence--based

### 2016-12-17 inst/manuscript/manuscript.rmd
Transition to use of bookdown for cross referencing of figures etc.; Switch from 'kable' to 'pandner' for table formatting, as text formatting in caption works flawlessly using it;

---
title: '`flippant` --- An `R` package for the automated analysis of fluorescence--based
  scramblase assays'
author: Richard J. Cotton ^†^, Birgit Ploier ^‡^, Michael A. Goren ^‡^, Anant K. Menon^‡^
  and Johannes Graumann ^†,§,^*
date: '`r format(Sys.Date(), "%B %d,%Y")`'
output:
  pdf_document:
    fig_caption: yes
  html_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
csl: biomed-central.csl
bibliography: flippant.bib
---

@@

# Abstract
**Background:** The lipid scrambling activity of protein extracts and purified
scramblases is typically measured using a fluorescence--based assay. While the
assay has yielded insight into the scramblase activity in crude membrane
preparations, functional validation of candidate scramblases, stoichiometry of
scramblase complexes as well as ATP--dependence of flippases, data analysis in
its context has remained a task involving many manual steps.

@@

# Keywords
scramblase; dithionite scramblase assay; R

# Background
Scramblases are proteins mediating the bidirectional, mass--action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes
[@waechter_flippases_2013;@pomorski_lipid_2006;@brunner_structural_2016].
Independent of energy  equivalents such as ATP they facilitate this "flipping"
by providing a mechanism by which polar lipid head groups can transition through
the hydrophobic environment of membranes, thus reducing the massive energetic
cost this process would carry otherwise [@kornberg_inside-outside_1971]. The
underlying molecular  mechanisms remain largely enigmatic [for a discussion
including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
Nonetheless, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta~2$--adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, one may speculate that lipid scrambling may, in
fact, be an intrinsic property of G--protein coupled receptors in general 
[@goren_constitutive_2014].

@@

# Availablility and Implementation
`flippant` has been implemented in `R` [@r_core_team_r:_2016] and integrated 
into the `Bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install-- and loadable as follows from 
within a functional `R` environment:

@@

# Input Data and Data Processing
To analyse raw data stemming from dithionite scramblase assays, `flippant` 
requires for each data point information on the path to the 
fluorimeter--generated spectrum, the amount of protein reconstituted into
proteoliposomes, the assay volume before and after addition of 
dithionite (to calculate a volume correction factor), the amount of lipids 
present (to calculate a protein to phospholipid ratio or PPR) etc. The user 
supplies this information in a spreadsheet (see below for supporting 
functionality).

@@

# Main Features
## Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program--compatible tab--delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.

@@

## Graphic Representation of Spectral Traces
In a first, graphical data analytic step of a scramblase assay--derived data 
set, fluorescent spectra or traces may be plotted in publication--appropriate
quality with acquisition time on the x-- and fluorescence intensity on the
y--axis using

@@

## Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart
of the analysis of dithionite scramblase assays, allowing for the visual
comparison of specific scramblase activity between interrogated experimental
series, which may, for example, be representative of protein preparations
including mutant protein forms or crude extracts with or without depletion of a
candidate scramblase.

@@

# Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G--protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by the recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved. However, the same set of experiments pointed at a
defect  in rhodopsin dimerization prior to insertion into liposomes, generating
the new hypothesis that the mutations may cause disease by interfering with the
highly ordered quarternary structure of rhodopsin in the retina 
[@ploier_dimerization_2016].

@@

To begin the analysis it is useful to plot the spectral traces of each 
experimental series (Figure 1). Note
that algorithmic alignment to the time of dithionite addition is employed
(default) as well as  custom trimming of the time axis to account for divergent
acquisition  times between experiments. PPR scaling is used with the default
factor of 0.65.

@@

The call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 2A.

Figure 2A emphasizes reproducibility between independent experiments
("Experiment 1" and "Experiment 2") on the four probed rhodopsins. Forgoing
reproducibility analysis in favor of a potentially more reliable result and
following more closely the analysis in Figure 4C--F by
@ploier_dimerization_2016, a single fit may be performed for the combined data
from all experiments as follows, resulting in Figure 2B:

@@

The graphical representation of the fit constant $\alpha$ in the PPR plots is
difficult to visually evaluate and easy access to the numeric values desirable.
Recapitulation of the calculations underlying PPR plot generation for Figure 2B
by `scramblase_assay_plot` in a call to `scramblase_assay_stats`, this is achieved
and produces tabular output represented in Table 1.

@@

```{r analysis_tabular_output, echo = FALSE}
ppr_results <- flippant::scramblase_assay_stats(
    template_file, split_by_experiment = FALSE) %>%
  dplyr::mutate(`Fit Constant` = round(`Fit Constant` * 10^4, 2)) %>%
  magrittr::set_names(c("Experimental Series", "Fit Constant ($\\times 10^4$)"))
ppr_results_names <- ppr_results %>%
  magrittr::extract2("Experimental Series") %>%
  to_string
ppr_results_alpha_print <- ppr_results %>%
  magrittr::extract2("Fit Constant ($\\times 10^4$)") %>%
  round(2) %>%
  to_string
knitr::kable(
  ppr_results,
  row.names = FALSE,
  caption = "**Tabular output of `flippant`--based reanalysis of a subset of the data in Figure 4C--F from @ploier_dimerization_2016.** An plain text--version of this output is produced by calling `scramblase_assay_stats`.")
```

@@

# Conclusions

```{r analysis_timing, echo = FALSE, cache = TRUE}

@@

# Declarations
## Ethics approval and consent to participate
Not applicable.

## Consent to publish
Not applicable.

## Availability of data and materials
The `R` package `flippant` provides all functionality as well as data presented
here along with extensive additional documentation and example code accessible
from within `R` in an integrated manner.

@@

## Competing interests
The authors declare that they have no competing interests.

## Funding
J.G., R.J.C. and the Proteomics Core at WCM--Q are supported by ``Biomedical 
Research Program'' funds at Weill Cornell Medicine -- Qatar, a program 
funded by Qatar Foundation. This study was made possible by NPRP grant
5--669--1--112 from the Qatar National Research Fund (a member of Qatar 
Foundation). The statements made herein are solely the responsibility of the
authors. B.P. was also supported by the Austrian Science Fund (FWF), project
J3686--B21.

## Authors' contributions
R.J.C and J.G. wrote the code, B.P., M.A.G. and A.K.M. guided the implementation
and edited the manuscript, which J.G. wrote. B.P. provided the exemplary assay
data. A.K.M. and J.G. supervised the work.

## Acknowledgements
Not applicable.

# Figure legends
```{r figure1-creation, echo=FALSE, fig.height = 4, fig.cap = "**`flippant`--based plotting of spectral traces underlying a subset of the data in Figure 4C--F from @ploier_dimerization_2016.** See text for command and details."}
# Plot spectral traces (Figure 1)
figure1_plot <- scramblase_assay_traces(
  template_file,
  time_max_sec = 350)
grid::rectGrob(
    width = grid::unit(0.1, "mm"),
    height = grid::unit(0.1, "mm")) %>%
  grid::grid.draw()
```

@@

# References

>>>>

---
title: '`flippant` --- An `R` package for the automated analysis of fluorescence--based
  scramblase assays'
author: Richard J. Cotton ^†^, Birgit Ploier ^‡^, Michael A. Goren ^‡^, Anant K. Menon^‡^
  and Johannes Graumann ^†,§,^*
date: '`r format(Sys.Date(), "%B %d,%Y")`'
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
  html_document:
    fig_caption: yes
  bookdown::word_document2:
    toc: no
    fig_caption: yes
csl: biomed-central.csl
bibliography: flippant.bib
---

@@

# Abstract {-}
**Background:** The lipid scrambling activity of protein extracts and purified
scramblases is typically measured using a fluorescence--based assay. While the
assay has yielded insight into the scramblase activity in crude membrane
preparations, functional validation of candidate scramblases, stoichiometry of
scramblase complexes as well as ATP--dependence of flippases, data analysis in
its context has remained a task involving many manual steps.

@@

# Keywords {-}
scramblase; dithionite scramblase assay; R

# Background {-}
Scramblases are proteins mediating the bidirectional, mass--action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes
[@waechter_flippases_2013;@pomorski_lipid_2006;@brunner_structural_2016].
Independent of energy  equivalents such as ATP they facilitate this "flipping"
by providing a mechanism by which polar lipid head groups can transition through
the hydrophobic environment of membranes, thus reducing the massive energetic
cost this process would carry otherwise [@kornberg_inside-outside_1971]. The
underlying molecular  mechanisms remain largely enigmatic [for a discussion
including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
Nonetheless, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta~2$--adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, one may speculate that lipid scrambling may, in
fact, be an intrinsic property of G--protein coupled receptors in general 
[@goren_constitutive_2014].

@@

# Availablility and Implementation {-}
`flippant` has been implemented in `R` [@r_core_team_r:_2016] and integrated 
into the `Bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install-- and loadable as follows from 
within a functional `R` environment:

@@

# Input Data and Data Processing {-}
To analyse raw data stemming from dithionite scramblase assays, `flippant` 
requires for each data point information on the path to the 
fluorimeter--generated spectrum, the amount of protein reconstituted into
proteoliposomes, the assay volume before and after addition of 
dithionite (to calculate a volume correction factor), the amount of lipids 
present (to calculate a protein to phospholipid ratio or PPR) etc. The user 
supplies this information in a spreadsheet (see below for supporting 
functionality).

@@

# Main Features {-}
## Template Generation for Data Input {-}
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program--compatible tab--delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.

@@

## Graphic Representation of Spectral Traces {-}
In a first, graphical data analytic step of a scramblase assay--derived data 
set, fluorescent spectra or traces may be plotted in publication--appropriate
quality with acquisition time on the x-- and fluorescence intensity on the
y--axis using

@@

## Scramblase Assay Analysis {-}
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart
of the analysis of dithionite scramblase assays, allowing for the visual
comparison of specific scramblase activity between interrogated experimental
series, which may, for example, be representative of protein preparations
including mutant protein forms or crude extracts with or without depletion of a
candidate scramblase.

@@

# Case Study {-}
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G--protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by the recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved. However, the same set of experiments pointed at a
defect  in rhodopsin dimerization prior to insertion into liposomes, generating
the new hypothesis that the mutations may cause disease by interfering with the
highly ordered quarternary structure of rhodopsin in the retina 
[@ploier_dimerization_2016].

@@

To begin the analysis it is useful to plot the spectral traces of each 
experimental series (Figure \@ref(fig:figure1-creation)). Note
that algorithmic alignment to the time of dithionite addition is employed
(default) as well as  custom trimming of the time axis to account for divergent
acquisition  times between experiments. PPR scaling is used with the default
factor of 0.65.

@@

The call results in the protein to phospholipid ratio (PPR) plot shown in
Figure \@ref(fig:figure2-creation)A.

Figure \@ref(fig:figure2-creation) A emphasizes reproducibility between
independent experiments ("Experiment 1" and "Experiment 2") on the four probed
rhodopsins. Forgoing reproducibility analysis in favor of a potentially more
reliable result and following more closely the analysis in Figure 4C--F by
@ploier_dimerization_2016, a single fit may be performed for the combined data
from all experiments as follows, resulting in 
Figure \@ref(fig:figure2-creation)B:

@@

The graphical representation of the fit constant $\alpha$ in the PPR plots is
difficult to visually evaluate and easy access to the numeric values desirable.
Recapitulation of the calculations underlying PPR plot generation for Figure \@ref(fig:figure2-creation)B by `scramblase_assay_plot` in a call to
`scramblase_assay_stats`, this is achieved and produces tabular output
represented in Table \@ref(tab:analysis-tabular-output).

@@

```{r analysis-tabular-output, echo = FALSE}
ppr_results <- flippant::scramblase_assay_stats(
    template_file, split_by_experiment = FALSE) %>%
  dplyr::mutate(`Fit Constant` = round(`Fit Constant` * 10^4, 2)) %>%
  magrittr::set_names(c("Experimental Series", "Fit Constant ($\\times 10^4$)"))
ppr_results_names <- ppr_results %>%
  magrittr::extract2("Experimental Series") %>%
  to_string
ppr_results_alpha_print <- ppr_results %>%
  magrittr::extract2("Fit Constant ($\\times 10^4$)") %>%
  round(2) %>%
  to_string
ppr_results %>%
  pander::pander(
    caption = "(\\#tab:analysis-tabular-output) **Tabular output of `flippant`--based reanalysis of a subset of the data in Figure 4C--F from @ploier_dimerization_2016.** An plain text--version of this output is produced by calling `scramblase_assay_stats`.")
```

@@

# Conclusions {-}
```{r analysis_timing, echo = FALSE, cache = TRUE}

@@

# Declarations {-}
## Ethics approval and consent to participate {-}
Not applicable.

## Consent to publish {-}
Not applicable.

## Availability of data and materials {-}
The `R` package `flippant` provides all functionality as well as data presented
here along with extensive additional documentation and example code accessible
from within `R` in an integrated manner.

@@

## Competing interests {-}
The authors declare that they have no competing interests.

## Funding {-}
J.G., R.J.C. and the Proteomics Core at WCM--Q are supported by ``Biomedical 
Research Program'' funds at Weill Cornell Medicine -- Qatar, a program 
funded by Qatar Foundation. This study was made possible by NPRP grant
5--669--1--112 from the Qatar National Research Fund (a member of Qatar 
Foundation). The statements made herein are solely the responsibility of the
authors. B.P. was also supported by the Austrian Science Fund (FWF), project
J3686--B21.

## Authors' contributions {-}
R.J.C and J.G. wrote the code, B.P., M.A.G. and A.K.M. guided the implementation
and edited the manuscript, which J.G. wrote. B.P. provided the exemplary assay
data. A.K.M. and J.G. supervised the work.

## Acknowledgements {-}
Not applicable.

# Figure legends {-}
```{r figure1-creation, echo=FALSE, fig.height = 4, fig.cap = "**`flippant`--based plotting of spectral traces underlying a subset of the data in Figure 4C--F from @ploier_dimerization_2016.** See text for command and details."}
# Plot spectral traces (Figure 1)
figure1_plot <- scramblase_assay_traces(
  template_file,
  time_max_sec = 350)
grid::rectGrob(
    width = grid::unit(0.1, "mm"),
    height = grid::unit(0.1, "mm")) %>%
  grid::grid.draw()
```

@@

# References {-}
